# IST 707 Data Analytics HOMEWORK 1

## Rshiny Dashboard Link: https://bhavishkumar.shinyapps.io/IST707HW1/

```{r}
setwd("C:/Users/bhavi/OneDrive/Desktop/SYR ADS/Sem 2/IST_707_Data_Analytics/HW")

getwd

emp_attr <- read.csv("employee_attrition.csv")
```


### Writing a function to obtain mode (most frequently occuring value)
*Match returns the first position of occurence of every element of 'x' in uniq*
*These position match indexes are tabulated which generates number of times each index is present*
*which.max provides the index of the maximum occuring match, i.e. the mode*
```{r}
getmode <- function(x) {       
  
  uniq <- unique(x)
  uniq[which.max(tabulate(match(x,uniq)))]
}
```

### Treating all missing values, by replacing with mean/median/mode

*Storing all missing values as NA*
*11 NAs in the table*

```{r}
emp_attr[emp_attr == ""] <-NA 
sum(is.na(emp_attr))   
```


*1 NA in the Gender column*
*Replace Gender NA with Mode of the column*
```{r}
sum(is.na(emp_attr$Gender))   
emp_attr$Gender[is.na(emp_attr$Gender)] <- getmode(emp_attr$Gender) 
```

*1 NA in the OverTime column*
*Replace OverTime NA with mode*
```{r}
emp_attr$OverTime[is.na(emp_attr$OverTime)] <- getmode(emp_attr$OverTime)
sum(is.na(emp_attr$OverTime))   
```

*2 NAs in DistanceFromHome*
*Replace NA with mean*
```{r}
emp_attr$DistanceFromHome[is.na(emp_attr$DistanceFromHome)] <- mean(emp_attr$DistanceFromHome, na.rm = TRUE) 
sum(is.na(emp_attr$DistanceFromHome))  
```

*1 NA in JobLevel column*
*Replace with mode*
```{r}
emp_attr$JobLevel[is.na(emp_attr$JobLevel)] <- getmode(emp_attr$JobLevel) 
sum(is.na(emp_attr$JobLevel)) 
```

  
*1 NA in percent salary hike*
*Replace NA with mean*
```{r}
emp_attr$PercentSalaryHike[is.na(emp_attr$PercentSalaryHike)] <- mean(emp_attr$PercentSalaryHike, na.rm = TRUE)
sum(is.na(emp_attr$PercentSalaryHike))
```


*1 NA in PerformanceRating*
*Replace NA with mode*
```{r}
emp_attr$PerformanceRating[is.na(emp_attr$PerformanceRating)] <- getmode(emp_attr$PerformanceRating)
sum(is.na(emp_attr$PerformanceRating))    
```


*1 NA in RelationshipSatisfaction*
*Replace NA with mode*
```{r}
emp_attr$RelationshipSatisfaction[is.na(emp_attr$RelationshipSatisfaction)] <- getmode(emp_attr$RelationshipSatisfaction)
sum(is.na(emp_attr$RelationshipSatisfaction)) 
```


*2 NA in Total Working Years*
*Replace NA with median*
```{r}
emp_attr$TotalWorkingYears[is.na(emp_attr$TotalWorkingYears)] <- median(emp_attr$TotalWorkingYears, na.rm = TRUE)
sum(is.na(emp_attr$TotalWorkingYears))
```


*1 NA in YearsSinceLastPromotion*                                                  
*Replace NA with median*
```{r}
emp_attr$YearsSinceLastPromotion[is.na(emp_attr$YearsSinceLastPromotion)] <- median(emp_attr$YearsSinceLastPromotion, na.rm = TRUE)
sum(is.na(emp_attr$YearsSinceLastPromotion))
```



```{r}
emp_attr_categorical_variables<-emp_attr
```


### HISTOGRAMS on numeric columns to identify Outliers and look at frequency distribution and spread

**HISTOGRAM of AGE column.Mean aproximately = Median, hence normal distribution**
```{r}
hist(emp_attr_categorical_variables$Age)
abline(v=mean(emp_attr_categorical_variables$Age),col = "red",lwd = 4)
abline(v=median(emp_attr_categorical_variables$Age),col = "blue", lwd = 2)
legend(x = "topright", 
       c("Mean", "Median"),
       col = c("red", "blue"),lwd = c(4,2))
```



**HISTOGRAM of Daily Rate column. ## Mean = Median, with Uniform distribution**
```{r}
hist(emp_attr_categorical_variables$DailyRate)
abline(v=mean(emp_attr_categorical_variables$DailyRate),col = "red",lwd = 4)
abline(v=median(emp_attr_categorical_variables$DailyRate),col = "blue", lwd = 2)
legend(x = "topright", 
       c("Mean", "Median"),
       col = c("red", "blue"),lwd = c(4,2))
```


**HISTOGRAM of DistanceFromHome column. Mean > Median, hence right skewed distribution**
```{r}
hist(emp_attr_categorical_variables$DistanceFromHome, breaks = 50)
abline(v=mean(emp_attr_categorical_variables$DistanceFromHome),col = "red",lwd = 4)
abline(v=median(emp_attr_categorical_variables$DistanceFromHome),col = "blue", lwd = 2)
legend(x = "topright", 
       c("Mean", "Median"),
       col = c("red", "blue"),lwd = c(4,2))
```


**HISTOGRAM of HourlyRate column. Mean = Median, with uniform distribution**
```{r}
hist(emp_attr_categorical_variables$HourlyRate)
abline(v=mean(emp_attr_categorical_variables$HourlyRate),col = "red",lwd = 4)
abline(v=median(emp_attr_categorical_variables$HourlyRate),col = "blue", lwd = 2)
legend(x = "topright", 
       c("Mean", "Median"),
       col = c("red", "blue"),lwd = c(4,2))
```


**HISTOGRAM of MonthlyIncome column. Mean > Median, Right Skewed with possibility of outliers**
```{r}
hist(emp_attr_categorical_variables$MonthlyIncome)
abline(v=mean(emp_attr_categorical_variables$MonthlyIncome),col = "red",lwd = 4)
abline(v=median(emp_attr_categorical_variables$MonthlyIncome),col = "blue", lwd = 2)
legend(x = "topright", 
       c("Mean", "Median"),
       col = c("red", "blue"),lwd = c(4,2))
```


**HISTOGRAM of MonthlyRate column. Mean = Median with Uniform Distribution.**
```{r}
hist(emp_attr_categorical_variables$MonthlyRate)
abline(v=mean(emp_attr_categorical_variables$MonthlyRate),col = "red",lwd = 4)
abline(v=median(emp_attr_categorical_variables$MonthlyRate),col = "blue", lwd = 2)
legend(x = "topright", 
       c("Mean", "Median"),
       col = c("red", "blue"),lwd = c(4,2))
```


**HISTOGRAM of NumCompaniesWorked column. Mean > Median with right skewed Distribution.**
```{r}
hist(emp_attr_categorical_variables$NumCompaniesWorked)
abline(v=mean(emp_attr_categorical_variables$NumCompaniesWorked),col = "red",lwd = 4)
abline(v=median(emp_attr_categorical_variables$NumCompaniesWorked),col = "blue", lwd = 2)
legend(x = "topright", 
       c("Mean", "Median"),
       col = c("red", "blue"),lwd = c(4,2))
```



**Histogram of YearsAtCompany column. Mean > Median with right skewed Distribution**
```{r}
hist(emp_attr_categorical_variables$YearsAtCompany)
abline(v=mean(emp_attr_categorical_variables$YearsAtCompany),col = "red",lwd = 4)
abline(v=median(emp_attr_categorical_variables$YearsAtCompany),col = "blue", lwd = 2)
legend(x = "topright", 
       c("Mean", "Median"),
       col = c("red", "blue"),lwd = c(4,2))
```

### Binning of numeric continuous variables to ordinal categorical variables for EDA & Association Rules creation

**Creating Age Groups**
```{r}
library(stringr)
emp_attr_categorical_variables$age_groups <-cut(emp_attr_categorical_variables$Age, breaks = c(quantile(emp_attr_categorical_variables$Age, probs = c(0,0.25,0.5,0.75,1))),
     labels = c(str_c(quantile(emp_attr_categorical_variables$Age,probs = 0),quantile(emp_attr_categorical_variables$Age,probs = 0.25),sep = " to "),str_c(quantile(emp_attr_categorical_variables$Age,probs = 0.25),quantile(emp_attr_categorical_variables$Age,probs = 0.5),sep = " to "),str_c(quantile(emp_attr_categorical_variables$Age,probs = 0.5),quantile(emp_attr_categorical_variables$Age,probs = 0.75),sep = " to "),str_c(quantile(emp_attr_categorical_variables$Age,probs = 0.75),quantile(emp_attr_categorical_variables$Age,probs = 1),sep = " to ")), right = FALSE, include.lowest=TRUE)
emp_attr_categorical_variables$age_groups<-as.factor(emp_attr_categorical_variables$age_groups)
unique(emp_attr_categorical_variables$age_groups)
```


**Creating Daily Rate Groups**
```{r}
emp_attr_categorical_variables$DailyRate_groups <-cut(emp_attr_categorical_variables$DailyRate, breaks = c(quantile(emp_attr_categorical_variables$DailyRate, probs = c(0,0.25,0.5,0.75,1))),
  labels = c(str_c(quantile(emp_attr_categorical_variables$DailyRate,probs = 0),quantile(emp_attr_categorical_variables$DailyRate,probs = 0.25),sep = " to "),str_c(quantile(emp_attr_categorical_variables$DailyRate,probs = 0.25),quantile(emp_attr_categorical_variables$DailyRate,probs = 0.5),sep = " to "),str_c(quantile(emp_attr_categorical_variables$DailyRate,probs = 0.5),quantile(emp_attr_categorical_variables$DailyRate,probs = 0.75),sep = " to "),str_c(quantile(emp_attr_categorical_variables$DailyRate,probs = 0.75),quantile(emp_attr_categorical_variables$DailyRate,probs = 1),sep = " to ")), right = FALSE, include.lowest=TRUE)
emp_attr_categorical_variables$DailyRate_groups<-as.factor(emp_attr_categorical_variables$DailyRate_groups)
unique(emp_attr_categorical_variables$DailyRate_groups)
```


**Creating DistanceFromHome Groups**
```{r}
emp_attr_categorical_variables$DistanceFromHome_groups <-cut(emp_attr_categorical_variables$DistanceFromHome, breaks = c(quantile(emp_attr_categorical_variables$DistanceFromHome, probs = c(0,0.25,0.5,0.75,1))),
labels = c(str_c(quantile(emp_attr_categorical_variables$DistanceFromHome,probs = 0),quantile(emp_attr_categorical_variables$DistanceFromHome,probs = 0.25),sep = " to "),str_c(quantile(emp_attr_categorical_variables$DistanceFromHome,probs = 0.25),quantile(emp_attr_categorical_variables$DistanceFromHome,probs = 0.5),sep = " to "),str_c(quantile(emp_attr_categorical_variables$DistanceFromHome,probs = 0.5),quantile(emp_attr_categorical_variables$DistanceFromHome,probs = 0.75),sep = " to "),str_c(quantile(emp_attr_categorical_variables$DistanceFromHome,probs = 0.75),quantile(emp_attr_categorical_variables$DistanceFromHome,probs = 0.99),sep = " to ")), right = FALSE, include.lowest=TRUE)
emp_attr_categorical_variables$DistanceFromHome_groups <- as.factor(emp_attr_categorical_variables$DistanceFromHome_groups)
unique(emp_attr_categorical_variables$DistanceFromHome_groups)
```

**Creating HourlyRate Groups**
```{r}
emp_attr_categorical_variables$HourlyRate_groups <- cut(emp_attr_categorical_variables$HourlyRate,breaks = c(quantile(emp_attr_categorical_variables$HourlyRate, probs = c(0,0.25,0.5,0.75,1))),
  labels = c(str_c(quantile(emp_attr_categorical_variables$HourlyRate,probs = 0),quantile(emp_attr_categorical_variables$HourlyRate,probs = 0.25),sep = " to "),str_c(quantile(emp_attr_categorical_variables$HourlyRate,probs = 0.25),quantile(emp_attr_categorical_variables$HourlyRate,probs = 0.5),sep = " to "),str_c(quantile(emp_attr_categorical_variables$HourlyRate,probs = 0.5),quantile(emp_attr_categorical_variables$HourlyRate,probs = 0.75),sep = " to "),str_c(quantile(emp_attr_categorical_variables$HourlyRate,probs = 0.75),quantile(emp_attr_categorical_variables$HourlyRate,probs = 1),sep = " to ")), right = FALSE,include.lowest=TRUE)
emp_attr_categorical_variables$HourlyRate_groups <- as.factor(emp_attr_categorical_variables$HourlyRate_groups)
unique(emp_attr_categorical_variables$HourlyRate_groups)
```


**Creating MonthlyIncome Groups**
```{r}
emp_attr_categorical_variables$MonthlyIncome_groups <- cut(emp_attr_categorical_variables$MonthlyIncome,breaks = c(quantile(emp_attr_categorical_variables$MonthlyIncome, probs = c(0,0.25,0.5,0.75,1))),
  labels = c(str_c(quantile(emp_attr_categorical_variables$MonthlyIncome,probs = 0),quantile(emp_attr_categorical_variables$MonthlyIncome,probs = 0.25),sep = " to "),str_c(quantile(emp_attr_categorical_variables$MonthlyIncome,probs = 0.25),quantile(emp_attr_categorical_variables$MonthlyIncome,probs = 0.5),sep = " to "),str_c(quantile(emp_attr_categorical_variables$MonthlyIncome,probs = 0.5),quantile(emp_attr_categorical_variables$MonthlyIncome,probs = 0.75),sep = " to "),str_c(quantile(emp_attr_categorical_variables$MonthlyIncome,probs = 0.75),quantile(emp_attr_categorical_variables$MonthlyIncome,probs = 1),sep = " to ")), right = FALSE,include.lowest=TRUE)
emp_attr_categorical_variables$MonthlyIncome_groups <- as.factor(emp_attr_categorical_variables$MonthlyIncome_groups)
unique(emp_attr_categorical_variables$MonthlyIncome_groups)
```

**Creating MonthlyRate Groups**
```{r}
emp_attr_categorical_variables$MonthlyRate_groups <- cut(emp_attr_categorical_variables$MonthlyRate,breaks = c(quantile(emp_attr_categorical_variables$MonthlyRate, probs = c(0,0.25,0.5,0.75,1))),
labels = c(str_c(quantile(emp_attr_categorical_variables$MonthlyRate,probs = 0),quantile(emp_attr_categorical_variables$MonthlyRate,probs = 0.25),sep = " to "),str_c(quantile(emp_attr_categorical_variables$MonthlyRate,probs = 0.25),quantile(emp_attr_categorical_variables$MonthlyRate,probs = 0.5),sep = " to "),str_c(quantile(emp_attr_categorical_variables$MonthlyRate,probs = 0.5),quantile(emp_attr_categorical_variables$MonthlyRate,probs = 0.75),sep = " to "),str_c(quantile(emp_attr_categorical_variables$MonthlyRate,probs = 0.75),quantile(emp_attr_categorical_variables$MonthlyRate,probs = 1),sep = " to ")), right = FALSE, include.lowest=TRUE)
emp_attr_categorical_variables$MonthlyRate_groups <- as.factor(emp_attr_categorical_variables$MonthlyRate_groups)
unique(emp_attr_categorical_variables$MonthlyRate_groups)
```

**Creating NumCompaniesWorked Groups**
```{r}
emp_attr_categorical_variables$NumCompaniesWorked_groups <- cut(emp_attr_categorical_variables$NumCompaniesWorked,breaks = c(quantile(emp_attr_categorical_variables$NumCompaniesWorked, probs = c(0,0.25,0.5,0.75,1))),
labels = c(str_c(quantile(emp_attr_categorical_variables$NumCompaniesWorked,probs = 0),quantile(emp_attr_categorical_variables$NumCompaniesWorked,probs = 0.25),sep = " to "),str_c(quantile(emp_attr_categorical_variables$NumCompaniesWorked,probs = 0.25),quantile(emp_attr_categorical_variables$NumCompaniesWorked,probs = 0.5),sep = " to "),str_c(quantile(emp_attr_categorical_variables$NumCompaniesWorked,probs = 0.5),quantile(emp_attr_categorical_variables$NumCompaniesWorked,probs = 0.75),sep = " to "),str_c(quantile(emp_attr_categorical_variables$NumCompaniesWorked,probs = 0.75),quantile(emp_attr_categorical_variables$NumCompaniesWorked,probs = 1),sep = " to ")), right = FALSE, include.lowest=TRUE)
emp_attr_categorical_variables$NumCompaniesWorked_groups <- as.factor(emp_attr_categorical_variables$NumCompaniesWorked_groups)
unique(emp_attr_categorical_variables$NumCompaniesWorked_groups)
```

**Creating PercentSalaryHike Groups**
```{r}
emp_attr_categorical_variables$PercentSalaryHike_groups <- cut(emp_attr_categorical_variables$PercentSalaryHike,breaks = c(quantile(emp_attr_categorical_variables$PercentSalaryHike, probs = c(0,0.25,0.5,0.75,1))),
labels = c(str_c(quantile(emp_attr_categorical_variables$PercentSalaryHike,probs = 0),quantile(emp_attr_categorical_variables$PercentSalaryHike,probs = 0.25),sep = " to "),str_c(quantile(emp_attr_categorical_variables$PercentSalaryHike,probs = 0.25),quantile(emp_attr_categorical_variables$PercentSalaryHike,probs = 0.5),sep = " to "),str_c(quantile(emp_attr_categorical_variables$PercentSalaryHike,probs = 0.5),quantile(emp_attr_categorical_variables$PercentSalaryHike,probs = 0.75),sep = " to "),str_c(quantile(emp_attr_categorical_variables$PercentSalaryHike,probs = 0.75),quantile(emp_attr_categorical_variables$PercentSalaryHike,probs = 1),sep = " to ")), right = FALSE, include.lowest=TRUE)
emp_attr_categorical_variables$PercentSalaryHike_groups <- as.factor(emp_attr_categorical_variables$PercentSalaryHike_groups)
unique(emp_attr_categorical_variables$PercentSalaryHike_groups)
```

**Creating TotalWorkingYears Groups**
```{r}
emp_attr_categorical_variables$TotalWorkingYears_groups <- cut(emp_attr_categorical_variables$TotalWorkingYears,breaks = c(quantile(emp_attr_categorical_variables$TotalWorkingYears, probs = c(0,0.25,0.5,0.75,1))),
labels = c(str_c(quantile(emp_attr_categorical_variables$TotalWorkingYears,probs = 0),quantile(emp_attr_categorical_variables$TotalWorkingYears,probs = 0.25),sep = " to "),str_c(quantile(emp_attr_categorical_variables$TotalWorkingYears,probs = 0.25),quantile(emp_attr_categorical_variables$TotalWorkingYears,probs = 0.5),sep = " to "),str_c(quantile(emp_attr_categorical_variables$TotalWorkingYears,probs = 0.5),quantile(emp_attr_categorical_variables$TotalWorkingYears,probs = 0.75),sep = " to "),str_c(quantile(emp_attr_categorical_variables$TotalWorkingYears,probs = 0.75),quantile(emp_attr_categorical_variables$TotalWorkingYears,probs = 0.99),sep = " to ")), right = FALSE, include.lowest=TRUE)
emp_attr_categorical_variables$TotalWorkingYears_groups <- as.factor(emp_attr_categorical_variables$TotalWorkingYears_groups)
unique(emp_attr_categorical_variables$TotalWorkingYears_groups)
```

**Creating YearsAtCompany Groups**
```{r}
emp_attr_categorical_variables$YearsAtCompany_groups <- cut(emp_attr_categorical_variables$YearsAtCompany,breaks = c(quantile(emp_attr_categorical_variables$YearsAtCompany, probs = c(0,0.25,0.5,0.75,1))),
labels = c(str_c(quantile(emp_attr_categorical_variables$YearsAtCompany,probs = 0),quantile(emp_attr_categorical_variables$YearsAtCompany,probs = 0.25),sep = " to "),str_c(quantile(emp_attr_categorical_variables$YearsAtCompany,probs = 0.25),quantile(emp_attr_categorical_variables$YearsAtCompany,probs = 0.5),sep = " to "),str_c(quantile(emp_attr_categorical_variables$YearsAtCompany,probs = 0.5),quantile(emp_attr_categorical_variables$YearsAtCompany,probs = 0.75),sep = " to "),str_c(quantile(emp_attr_categorical_variables$YearsAtCompany,probs = 0.75),quantile(emp_attr_categorical_variables$YearsAtCompany,probs = 1),sep = " to ")), right = FALSE, include.lowest=TRUE)
emp_attr_categorical_variables$YearsAtCompany_groups <- as.factor(emp_attr_categorical_variables$YearsAtCompany_groups)
unique(emp_attr_categorical_variables$YearsAtCompany_groups)
```

**Creating YearsInCurrentRole Groups**
```{r}
emp_attr_categorical_variables$YearsInCurrentRole_groups <- cut(emp_attr_categorical_variables$YearsInCurrentRole,breaks = c(quantile(emp_attr_categorical_variables$YearsInCurrentRole, probs = c(0,0.25,0.5,0.75,1))),
labels = c(str_c(quantile(emp_attr_categorical_variables$YearsInCurrentRole,probs = 0),quantile(emp_attr_categorical_variables$YearsInCurrentRole,probs = 0.25),sep = " to "),str_c(quantile(emp_attr_categorical_variables$YearsInCurrentRole,probs = 0.25),quantile(emp_attr_categorical_variables$YearsInCurrentRole,probs = 0.5),sep = " to "),str_c(quantile(emp_attr_categorical_variables$YearsInCurrentRole,probs = 0.5),quantile(emp_attr_categorical_variables$YearsInCurrentRole,probs = 0.75),sep = " to "),str_c(quantile(emp_attr_categorical_variables$YearsInCurrentRole,probs = 0.75),quantile(emp_attr_categorical_variables$YearsInCurrentRole,probs = 1),sep = " to ")), right = FALSE, include.lowest=TRUE)
emp_attr_categorical_variables$YearsInCurrentRole_groups <- as.factor(emp_attr_categorical_variables$YearsInCurrentRole_groups)
unique(emp_attr_categorical_variables$YearsInCurrentRole_groups)
```


**Creating YearsSinceLastPromotion Groups**
```{r}
emp_attr_categorical_variables$YearsSinceLastPromotion_groups <- cut(emp_attr_categorical_variables$YearsSinceLastPromotion,breaks = c(quantile(emp_attr_categorical_variables$YearsSinceLastPromotion, probs = c(0,0.5,0.75,1))),
labels = c(str_c(quantile(emp_attr_categorical_variables$YearsSinceLastPromotion,probs = 0),quantile(emp_attr_categorical_variables$YearsSinceLastPromotion,probs = 0.5),sep = " to "),str_c(quantile(emp_attr_categorical_variables$YearsSinceLastPromotion,probs = 0.5),quantile(emp_attr_categorical_variables$YearsSinceLastPromotion,probs = 0.75),sep = " to "),str_c(quantile(emp_attr_categorical_variables$YearsSinceLastPromotion,probs = 0.75),quantile(emp_attr_categorical_variables$YearsSinceLastPromotion,probs = 1),sep = " to ")), right = FALSE, include.lowest=TRUE)
emp_attr_categorical_variables$YearsSinceLastPromotion_groups <- as.factor(emp_attr_categorical_variables$YearsSinceLastPromotion_groups)
unique(emp_attr_categorical_variables$YearsSinceLastPromotion_groups)
```


**Creating YearsWithCurrManager Groups**
```{r}
emp_attr_categorical_variables$YearsWithCurrManager_groups <- cut(emp_attr_categorical_variables$YearsWithCurrManager,breaks = c(quantile(emp_attr_categorical_variables$YearsWithCurrManager, probs = c(0,0.25,0.5,0.75,1))),
labels = c(str_c(quantile(emp_attr_categorical_variables$YearsWithCurrManager,probs = 0),quantile(emp_attr_categorical_variables$YearsWithCurrManager,probs = 0.25),sep = " to "),str_c(quantile(emp_attr_categorical_variables$YearsWithCurrManager,probs = 0.25),quantile(emp_attr_categorical_variables$YearsWithCurrManager,probs = 0.5),sep = " to "),str_c(quantile(emp_attr_categorical_variables$YearsWithCurrManager,probs = 0.5),quantile(emp_attr_categorical_variables$YearsWithCurrManager,probs = 0.75),sep = " to "),str_c(quantile(emp_attr_categorical_variables$YearsWithCurrManager,probs = 0.75),quantile(emp_attr_categorical_variables$YearsWithCurrManager,probs = 0.99),sep = " to ")), right = FALSE, include.lowest=TRUE)
emp_attr_categorical_variables$YearsWithCurrManager_groups <- as.factor(emp_attr_categorical_variables$YearsWithCurrManager_groups)
unique(emp_attr_categorical_variables$YearsWithCurrManager_groups)
```


*CAN EXCLUDE EmployeeNumber, EmployeeCount, Over18 & StandardHours variables as they have 0 variance*
```{r}
emp_attr_categorical_variables$EmployeeNumber <- NULL
emp_attr_categorical_variables$EmployeeCount <- NULL
emp_attr_categorical_variables$Over18 <- NULL
emp_attr_categorical_variables$StandardHours <- NULL
```



### PERFORM EDA to identify interesting trends in data
```{r}
library(dplyr)
library(tidyverse)
library(plotly)
library(ggplot2)
```


### HYPOTHESIS
### 1. Attrition Rate is very low amongst married/divorced employees and very high amongst single employees
```{r}
df_marital_status <- emp_attr_categorical_variables %>%               
  group_by(MaritalStatus,Attrition) %>%
  summarise(count = n()) %>%
    spread(Attrition,count) %>%
    ungroup %>%
    transmute(MaritalStatus=MaritalStatus,attr_rate = Yes/(Yes+No))
  df_marital_status
  
  colnames(df_marital_status) <- c("Marital_Status","Attrition_Rate")
marital_status_attrition_plot <- ggplot(df_marital_status,aes(x=Marital_Status,y=Attrition_Rate))+
  geom_col(show.legend=TRUE,position = "dodge")+
  xlab("Marital Status")+ ylab("Attrition Rate")+
  ggtitle("Employee Attrition Rate by Marital Status")
marital_status_attrition_plot
```


  
### 2. Attrition Rate is very low amongst Non Travellers and higher amongst frequent travellers
```{r}
df_travel <- emp_attr_categorical_variables %>%             
  group_by(BusinessTravel,Attrition) %>%
  summarise(count = n()) %>%
  spread(Attrition,count) %>%
  ungroup %>%
  transmute(BusinessTravel=BusinessTravel,attr_rate = Yes/(Yes+No))
df_travel
colnames(df_travel) <- c("BusinessTravel","Attrition_Rate")
business_travel_attrition_plot <- ggplot(df_travel,aes(x=BusinessTravel,y=Attrition_Rate))+
  geom_col(show.legend=TRUE,position = "dodge")+
  xlab("Travel Rate")+ ylab("Attrition Rate")+
  ggtitle("Employee Attrition Rate by Travel Rate")
business_travel_attrition_plot
```


### 3. Attrition rate is lower amongst employees living near by and higher amongst employees staying far off
```{r}
df_distance <- emp_attr_categorical_variables %>%             
  group_by(DistanceFromHome_groups,Attrition) %>%
  summarise(count = n()) %>%
  spread(Attrition,count) %>%
  ungroup %>%
  transmute(DistanceFromHome_groups=DistanceFromHome_groups,attr_rate = Yes/(Yes+No))
df_distance
colnames(df_distance) <- c("DistanceFromHome_groups","Attrition_Rate")
distance_attrition_plot <- ggplot(df_distance,aes(x=DistanceFromHome_groups,y=Attrition_Rate))+
  geom_col(show.legend=TRUE,position = "dodge")+
  xlab("Distance from Home groups")+ ylab("Attrition Rate")+
  ggtitle("Employee Attrition Rate by distance from home groups")
distance_attrition_plot
```


### 4. Attrition Rate is lower amongst employees with high environment satisfaction and is higher amongst employees with low environment satisfaction
```{r}
df_envt <- emp_attr_categorical_variables %>%             
  group_by(EnvironmentSatisfaction,Attrition) %>%
  summarise(count = n()) %>%
  spread(Attrition,count) %>%
  ungroup %>%
  transmute(EnvironmentSatisfaction=EnvironmentSatisfaction,attr_rate = Yes/(Yes+No))
df_envt
colnames(df_envt) <- c("EnvironmentSatisfaction","Attrition_Rate")
envt_attrition_plot <- ggplot(df_envt,aes(x=EnvironmentSatisfaction,y=Attrition_Rate))+
  geom_col(show.legend=TRUE,position = "dodge")+
  xlab("Environment Satisfaction")+ ylab("Attrition Rate")+
  ggtitle("Employee Attrition Rate by environment satisfaction")
envt_attrition_plot
```


### 5. Attrition Rate is lower amongst high Monthly Income groups
```{r}
df_monthly_income <- emp_attr_categorical_variables %>%             
  group_by(MonthlyIncome_groups,Attrition) %>%
  summarise(count = n()) %>%
  spread(Attrition,count) %>%
  ungroup %>%
  transmute(MonthlyIncome_groups=MonthlyIncome_groups,attr_rate = Yes/(Yes+No))
df_monthly_income
colnames(df_monthly_income) <- c("MonthlyIncome_groups","Attrition_Rate")
monthly_income_attrition_plot <- ggplot(df_monthly_income,aes(x=MonthlyIncome_groups,y=Attrition_Rate))+
  geom_col(show.legend=TRUE,position = "dodge")+
  xlab("Monthly Income Groups")+ ylab("Attrition Rate")+
  ggtitle("Employee Attrition Rate by monthly income groups")
monthly_income_attrition_plot
```


### 6. Attrition Rate is higher amongst employees who have worked for fewer years at the company
```{r}
  df_yearsatcompany <- emp_attr_categorical_variables %>%             
  group_by(YearsAtCompany_groups,Attrition) %>%
  summarise(count = n()) %>%
  spread(Attrition,count) %>%
  ungroup %>%
  transmute(YearsAtCompany_groups=YearsAtCompany_groups,attr_rate = Yes/(Yes+No))
  df_yearsatcompany
  colnames(df_yearsatcompany) <- c("YearsAtCompany_groups","Attrition_Rate")
  yearsatcompany_plot <- ggplot(df_yearsatcompany,aes(x=YearsAtCompany_groups,y=Attrition_Rate))+
  geom_col(show.legend=TRUE,position = "dodge")+
  xlab("Years at company Groups")+ ylab("Attrition Rate")+
  ggtitle("Employee Attrition Rate by years at company groups")
  yearsatcompany_plot
```



### 7. Attrition Rate is higher amongst employees with lower job levels
```{r}
df_joblevel <- emp_attr_categorical_variables %>%             
  group_by(JobLevel,Attrition) %>%
  summarise(count = n()) %>%
  spread(Attrition,count) %>%
  ungroup %>%
  transmute(JobLevel=JobLevel,attr_rate = Yes/(Yes+No))
df_joblevel
colnames(df_joblevel) <- c("JobLevel","Attrition_Rate")
JobLevel_plot <- ggplot(df_joblevel,aes(x=JobLevel,y=Attrition_Rate))+
  geom_col(show.legend=TRUE,position = "dodge")+
  xlab("Job Level")+ ylab("Attrition Rate")+
  ggtitle("Employee Attrition Rate by Job Level")
JobLevel_plot
```


### 8. Attrition Rate is higher amongst lower age group employees
```{r}
df_age_groups <- emp_attr_categorical_variables %>%             
  group_by(age_groups,Attrition) %>%
  summarise(count = n()) %>%
  spread(Attrition,count) %>%
  ungroup %>%
  transmute(age_groups=age_groups,attr_rate = Yes/(Yes+No))
df_age_groups
colnames(df_age_groups) <- c("AgeGroups","Attrition_Rate")
age_group_plot <- ggplot(df_age_groups,aes(x=AgeGroups,y=Attrition_Rate))+
  geom_col(show.legend=TRUE,position = "dodge")+
  xlab("Age Groups")+ ylab("Attrition Rate")+
  ggtitle("Employee Attrition Rate by Age Groups")
age_group_plot
```


### Removing Continuous numeric variables from dataframe for running Association Rules Mining
```{r}
emp_attr_arm <- emp_attr_categorical_variables
emp_attr_arm$Age<-NULL
emp_attr_arm$DailyRate<-NULL
emp_attr_arm$DistanceFromHome<-NULL
emp_attr_arm$HourlyRate<-NULL
emp_attr_arm$MonthlyIncome<-NULL
emp_attr_arm$MonthlyRate<-NULL
emp_attr_arm$NumCompaniesWorked<-NULL
emp_attr_arm$PercentSalaryHike<-NULL
emp_attr_arm$TotalWorkingYears<-NULL
emp_attr_arm$YearsAtCompany<-NULL
emp_attr_arm$YearsInCurrentRole<-NULL
emp_attr_arm$YearsSinceLastPromotion<-NULL
emp_attr_arm$YearsWithCurrManager<-NULL
emp_attr_arm$Education <-as.factor(emp_attr_arm$Education)
emp_attr_arm$EnvironmentSatisfaction <- as.factor(emp_attr_arm$EnvironmentSatisfaction)
emp_attr_arm$JobInvolvement <- as.factor(emp_attr_arm$JobInvolvement)
emp_attr_arm$JobLevel <- as.factor(emp_attr_arm$JobLevel)
emp_attr_arm$JobSatisfaction <- as.factor(emp_attr_arm$JobSatisfaction)
emp_attr_arm$PerformanceRating <- as.factor(emp_attr_arm$PerformanceRating)
emp_attr_arm$RelationshipSatisfaction <- as.factor(emp_attr_arm$RelationshipSatisfaction)
emp_attr_arm$StockOptionLevel <- as.factor(emp_attr_arm$StockOptionLevel)
emp_attr_arm$TrainingTimesLastYear <- as.factor(emp_attr_arm$TrainingTimesLastYear)
emp_attr_arm$WorkLifeBalance <- as.factor(emp_attr_arm$WorkLifeBalance)
```


## ASSOCIATION RULES MINING, using APRIORI
**CREATING TRANSACTIONS MATRIX**
```{r}
#install.packages("arules")
library(arules)  

#install.packages("arulesViz")
library(arulesViz)

emp_attr_armX <- as(emp_attr_arm,"transactions")
```


### BASELINE MODEL
**BASELINE MODEL RULES for ATTRITION "NO", i.e. employees who have not attritioned**
```{r}
ruleset_attrition_no_baseline <- apriori(emp_attr_armX,                                                
                   appearance = list(default = "lhs", rhs=("Attrition=No")))
```


**INSPECT RULES for Attrition = 'No' Baseline model**
```{r}
inspect(head(sort (ruleset_attrition_no_baseline, by="lift", decreasing=TRUE),5))
```


**BASELINE MODEL RULES for ATTRITION "YES", i.e. employees who have attritioned**
```{r}
ruleset_attrition_yes_baseline <- apriori(emp_attr_armX,                                                
                   appearance = list(default = "lhs", rhs=("Attrition=Yes")))
```


**INSPECT RULES for Attrition = 'YES' Baseline model**
*No rules returned for baseline model*
```{r}
inspect(head(sort (ruleset_attrition_yes_baseline, by="lift", decreasing=TRUE),5))
```


### FINE TUNED MODELs by adjusting HYPERPARAMETERS
**RULES for ATTRITION "NO", i.e. employees who have not attritioned**
```{r}
ruleset_attrition_no <- apriori(emp_attr_armX,                                                
                   parameter = list(support = 0.35, confidence = 0.7, minlen = 3, maxlen = 15),      
                   appearance = list(default = "lhs", rhs=("Attrition=No")))
```


**RULES for ATTRITION "Yes", i.e. employees who have attritioned**
```{r}
ruleset_attrition_yes <- apriori(emp_attr_armX,                                                
                                parameter = list(support = 0.03, confidence = 0.6,  minlen = 3, maxlen = 15),      
                                appearance = list(default = "lhs", rhs=("Attrition=Yes")))
```

*By Fine Tuning the Hyperparameters, we can filter out the best rules, based on confidence and lift. Overfitting and Underfitting does not exist for Association Rules Mining, as it is an unsupervised learning algorithm. High Confidence implies, that number of times RHS has occured, given LHS has occured is high. Moreover, a high Lift along with high confidence makes the rule significant and interesting because, it not only tells us that out of all the times, LHS has occured, even RHS has occured, but also that, whenever RHS has occured, LHS has occured with it. i.e. High Confidence and High Lift means that product of support of RHS and support of LHS independently  is lower than support of (LHS and RHS) together*

#### Using the inspect() command, review the "ruleset" for Attrition = No, i.e. people who stay
*These are the top 5 most interesting and significant rules, because they are having the highest lift. Hence these rules can predict people who stay*
*Since lift is > 1, the LHS & RHS are positively correlated.*
```{r}
inspect(head(sort (ruleset_attrition_no, by="lift", decreasing=TRUE),5))
```


**Using the inspect() command, review the "ruleset" for Attrition = Yes, i.e. people who leave**
*These are the top 5 most interesting and significant rules, because they are having the highest lift. Hence these rules can predict people who leave*
*Since lift is > 1, the LHS & RHS are positively correlated.*
```{r}
inspect(head(sort (ruleset_attrition_yes, by="lift", decreasing=TRUE),5))
```

#### Experiment with the interactive ruleset interface by running the inspectDT() command
**For Attrition No**
```{r}
ruleset_attrition_no_sort <- sort (ruleset_attrition_no, by="lift", decreasing=TRUE)
inspectDT(ruleset_attrition_no_sort)
```

**For Attrition Yes**
```{r}
ruleset_attrition_yes_sort <- sort (ruleset_attrition_yes, by="lift", decreasing=TRUE)
inspectDT(ruleset_attrition_yes_sort)
```